Supporting Information for "Can we distinguish model of selective interactions using linkage disequilibrium?

### 2021-03-24 Manuscript.Rmd
ready for last pass of edits

(ref:fig1cap) \textbf{Additive selection and epistasis.} (A and B) The
distribution at stationarity of $AB$ haplotype counts in a sample size of 30,
in which we observe 10 $A$ alleles at the left locus, and 10 $B$ alleles at the
right locus. (C) The decay of normalized squared LD ($\sigma_d^2$) with
recombination distance ($\rho=4N_er$) for pairs of variants with a given
selection coefficient. The dashed line is the neutral expectation from
@Ohta1971-yd, given in Equation \@ref(eq:ohta). (D) The decay of
$\sigma_d^1=\E[D]/ \E[p(1-p)q(1-q)]$ for pairs of variants. The negative LD
effect of Hill-Robertson interference is strongest for intermediate selection,
with $\gamma\sim-1$ to $-5$, where $\gamma=4N_es$. **[The text does not refer
to specific panels, 1c should give the equation for $\sigma_d^2$, and not sure
what we get from 1a and 1b. Define rho in caption, and place other parameters
in parentheses]** **[Move a, b to supp, consolidate c and d with fig 2]** And
from figure 2 previously ...(A, B) Synergistic epistasis decreases
$\sigma_d^2$, while antagonistic epistasis increases $\sigma_d^2$, though both
are decreased in comparison to neutral expectations. (C, D) Epistasis expected
signed LD away from zero, an effect which can extend over large recombination
distances. Here, $\gamma$ is the selection coefficient for derived alleles $A$
and $B$, assuming selection is equal at the two loci. $\epsilon=0$ is the case
of pure additive selection with no epistasis, and $\gamma_{AB}=(\gamma_A +
\gamma_B)(1+\epsilon)$. Dashed lines are expectations under neutrality for both
$\sigma_d^2$ and $\sigma_d^1$. **[More explicitly define that positive e is
synergistic, negative is antagonistic, and define formula in the methods]**
**[caption for C, D reads a little trailing]**

@@

With epistasis, mean signed LD is large for both weakly and strongly selected
variants, with sign depending on the direction of epistatic interactions
(Figure \@ref(fig:fig1)). Synergistic epistasis results in negative LD while
antagonistic epistasis results in positive LD, and large positive LD can occur
even when the epistatic effect is relatively weak **[example in supp]**.
Epistasis-induced LD can extend over long distances, especially for strongly
deleterious mutations. For example, in Figure \@ref(fig:fig1)F even moderately
deleterious mutations with population-size-scaled selection coefficients of
$\gamma=-10$ show large mean LD that extends well beyond 100 kilobases. More
strongly deleterious interacting mutations are expected to show large signed LD
over much larger recombination distances.

@@

(ref:fig2cap) **Dominance effects.** (A-B) Pairs of recessive mutations
($h<0.5$) have higher squared LD than additive mutations, while pairs of
dominant alleles have lower expected $D^2$. (C-D) Pairs of recessive, strongly
deleterious mutations can have large negative LD, although this correlation
decays more rapidly with larger recombination than LD under epistasis (Figure
\@ref(fig:fig1)D and F). Notably, there is no simple relationship between strength of
selection and dominance, particularly when $s$ is $O(1/N)$. **[Epistasis is
zero in this setup, and rearrange panels to match figure 1]**

@@

**[Struggled a bit with this section, make clearer and easier to parse]**

While dominance and epistasis are well-characterized models of selection, a
large family of selection models can be specified by assigning unique
selection coefficients to possible diploid pairs of haplotypes. If we assume
the diploid genotype that is homozygous for the ancestral allele ($ab/ab$) has
fitness 1, then there are nine other possible diploid two-locus genotypes that
could be given unique fitnesses (Table \@ref(tab:selmodels)), noting that
$AB/ab$ and $Ab/aB$ genotypes can have differing fitness effects.

The case with $s_{AB/ab}\not=s_{Ab/aB}$ can arise in a scenario where a
mutation at either locus within a haplotype impacts some functional region or
element, but a diploid individual carrying at least one copy that is free of
mutations has minimal fitness loss. In this "gene-based dominance" scenario
[e.g., @Sanjak2017-zw], an $AB/ab$ genotype has higher fitness than an $Ab/aB$
type (a simple implementation of this model is given in the final column in
Table \@ref(tab:selmodels)). Such a gene-based fitness model gives similar
expected positive signed LD to the model of antagonistic epistasis (compare
Figure \@ref(fig:fig3)E-F to G-H), although the interpretation of those two
models can differ. It shows that with a highly parameterized space of possible
general diploid selection models, multiple models with different biological
interpretations can give similar patterns of expected $D$.

@@

Synonymous and missense mutations show similar levels of positive LD when
considering pairs of mutations within the same gene. Loss-of-function mutations
instead have more negative LD, suggesting differing modes of selective
interactions for loss-of-function and missense mutations (Figure
\@ref(fig:fig4)). Within each population, large measurement noise gives 95\%
confidence intervals that overlap with zero across mutation class, although the
pattern is remarkably consistent across African, European, and East Asian
populations in the Thousand Genomes dataset. Comparing mean LD within genes in
African and Eurasian populations, LD in Eurasian populations is somewhat larger
on average, that is, more positive for missense mutations and more negative for
loss-of-function mutations. This is in agreement with differences in
expectations between populations that have or have not gone through a
bottleneck in their recent past (Figure \@ref(fig:fig3)). For pairs of missense
mutations in different protein-coding genes, signed LD decays to zero fairly
rapidly and is lower than signed LD within genes even for mutations that are
linked (distance less that 50 kb).

@@

**[Mashaal points out that simulations with structure leading to inbreeding, in which subpopulations have differing load, can lead to positive LD]**

In a single randomly mating population, neutral mutations are expected to have
average signed LD of zero, but across all populations analyzed here, LD between
synonymous mutations is positive. While selection on some subset of synonymous
variants is possible, it should be much weaker on average than between missense
mutations, and any interference between selected synonymous variants should
lead to negative LD. @Sandler2020-vk proposed that spatial population structure
may be responsible for the increase of LD for both synonymous and missense
variants. They argued that every natural population lives in some geography
that will necessarily induce some structure, and that migration, admixture, or
spatial structure could increase average signed LD above zero.

Under standard population genetic assumptions of constant mutation and
recombination rates and independent mutation events, however, migration and
admixture do not result in nonzero $\E[D]$. Rather, for a set of mutations with
*given* differences in allele frequencies between source populations, $D$ can
be nonzero in an admixed population between those particular mutations even
when $D$ is zero in the source populations [@Cavalli-Sforza1971-jv]. But this
result relies on conditioning on directional differences in allele frequencies
between populations. Expected mean allele frequency difference in the source
populations is zero ($\E[(p_1-p_2)(q_1-q_2)] = 0$), as differences are equally
likely to be positive or negative, and resulting $\E[D]$ is still zero when
taking genome-wide averages. Furthermore, @Sandler2020-vk used forward
simulations with an explicit spatial model and a multiplicative fitness
function to explore the joint effect of structure and interference, and the
positive LD they observed under these scenarios extends over much larger
distances than observed in the data. Instead, positive LD between synonymous
mutations decays to zero within just a few hundred base (Figures
\@ref(fig:lddecay) and \@ref(fig:lddecaydomain)).

@@

**[Need to point out that confidence intervals are wide, and while the signal
is consistent across populations, shared evolutionary history makes formal
testing of significance difficult.]**

In exploring or inferring multi-locus selective models, the number of plausible
selection scenarios becomes quite large as we relax the strict assumptions of
additivity and multiplicative fitness effects. The inclusion of dominance
effects, epistasis, or other interactions leads to a rapid increase in the
number of parameters to consider. This makes performing forward simulations
that span the range of all such selective interaction scenarios burdensome.

@@

To include selection, we consider a model where we draw lineages uniformly from
the previous generation, but keep lineages with probability proportional to
their fitness. In the absence of dominance, selection reduces to a haploid
model, with acceptance and rejection probabilities depending on the fitnesses
of each haploid copy, where haplotype $Ab$ has fitness $1 + s_{A}$, $aB$ has
fitness $1 + s_{B}$, and $AB$ has fitness $1 + s_{AB}$. We assume the doubly
ancestral haplotype $ab$ has fitness $1$, so fitnesses are relative to that of
$ab$ haplotypes. The standard multiplicative fitness function assumes that
$s_{AB} \approx s_{A} + s_{B}$, and the effect of epistasis can be written as
$s_{AB} = (s_{A} + s_{B}) (1 + \epsilon)$, so that $\epsilon > 0$ implies
synergistic epistasis, while $\epsilon < 0$ implies antagonistic epistasis.

>>>>

(ref:fig1cap) \textbf{Additive selection and epistasis.} Left panels show
expectations for the decay of $\sigma_d^2 = \E[D^2] / \E[p(1-p)q(1-q)]$ with
recombination distance, and right panels show expectations for $\sigma_d^1 =
\E[D] / \E[p(1-p)q(1-q)]$. The scaled recombination rate is given by
$\rho=4N_er$, where $r$ is the probability of recombination between two loci.
All expectations are at stationarity, and dashed lines are expectations under
neutrality (Equation \@ref(eq:ohta)). (A-B) Additive selection, with equal
selection strength at both loci. Stronger selection reduces $\sigma_d^2$ (A),
and in (B), we see that the negative LD effect of Hill-Robertson interference
is strongest for intermediate selection ($\gamma \approx -1$ to $-5$, where
$\gamma=4N_es$). (C-F) Epistasis with additive selection. Synergistic epistasis
(\(\epsilon < 0\)) decreases $\sigma_d^2$ while antagonistic epistasis
(\(\epsilon > 0\)) increases it compared to expectations under a multiplicative
fitness model ($\epsilon=0$). Epistasis also pushes signed LD ($\sigma_d^1$)
away from zero, with synergistic epistasis decreasing LD for negatively
selected mutations and antagonistic epistasis increasing LD. The deviation of
$\sigma_d^1$ from zero under epistasis can extend for long recombination
distances, particularly for stronger selection.

@@

With epistasis, mean signed LD is large for both weakly and strongly selected
variants, with sign depending on the direction of epistatic interactions
(Figure \@ref(fig:fig1)). Synergistic epistasis results in negative LD while
antagonistic epistasis results in positive LD, and large positive LD can occur
even when the epistatic effect is relatively weak. Epistasis-induced LD can
extend over long distances, especially for strongly deleterious mutations. For
example, in Figure \@ref(fig:fig1)F even moderately deleterious mutations with
population-size-scaled selection coefficients of $\gamma=-10$ show large mean
LD that extends well beyond 100 kilobases. More strongly deleterious
interacting mutations are expected to show large signed LD over much larger
recombination distances.

@@

(ref:fig2cap) **Dominance effects.** (A and C) Pairs of recessive mutations
($h<0.5$) have higher squared LD than additive mutations, while pairs of
dominant mutations have lower expected $D^2$. (B and D) Pairs of recessive
deleterious mutations can have large negative LD, although this correlation
decays more rapidly with larger recombination than LD under epistasis (Figure
\@ref(fig:fig1)D and F). Notably, there is no simple relationship between
strength of selection and dominance, particularly when $s$ is $O(1/N)$. Note
that this setup does not include epistatic effects (i.e., $\epsilon=0$).

@@

While dominance and epistasis are well-characterized models of selection, a
large family of selection models can be specified by assigning unique
selection coefficients to each possible diploid pairs of haplotypes. If we assume
the diploid genotype that is homozygous for the ancestral alleles ($ab/ab$) has
fitness 1, then there are nine other possible diploid two-locus genotypes that
could be given unique fitnesses (Table \@ref(tab:selmodels)), noting that
$AB/ab$ and $Ab/aB$ genotypes can have differing fitness effects.

The case with $s_{AB/ab}\not=s_{Ab/aB}$ can arise in a scenario where a
mutation at either locus within a haplotype impacts some functional region or
element, but a diploid individual carrying at least one copy that is free of
mutations has minimal fitness loss. In this "gene-based dominance" scenario
[e.g., @Sanjak2017-zw], an $AB/ab$ genotype has higher fitness than an $Ab/aB$
type (a simple implementation of this model is given in the final column in
Table \@ref(tab:selmodels)). Such a gene-based fitness model gives similar
expected positive signed LD to the model of antagonistic epistasis (compare
Figure \@ref(fig:fig3)E-F to G-H), although the interpretation of those two
models can differ. With a highly parameterized space of possible general
diploid selection models, multiple models with different biological
interpretations can give similar patterns of expected signed LD.

@@

Synonymous and missense mutations show similar levels of positive LD when
considering pairs of mutations within the same gene. Loss-of-function mutations
instead have more negative LD, suggesting differing modes of selective
interactions for loss-of-function and missense mutations (Figure
\@ref(fig:fig4)). Within each population, measurement noise gives 95\%
confidence intervals that overlap with zero across mutation class, although the
pattern is remarkably consistent across African, European, and East Asian
populations in the Thousand Genomes dataset. Comparing mean LD within genes in
African and Eurasian populations, LD in Eurasian populations is somewhat larger
on average, that is, more positive for missense mutations and more negative for
loss-of-function mutations. This is in agreement with differences in
expectations between populations that have or have not gone through a
bottleneck in their recent past (Figure \@ref(fig:fig3)). For pairs of missense
mutations in different protein-coding genes, signed LD decays to zero fairly
rapidly and is lower than signed LD within genes even for mutations that are
linked (distance less that 50 kb).

@@

In a single randomly mating population, neutral mutations are expected to have
average signed LD of zero, but across all populations analyzed here, LD between
synonymous mutations is positive. While selection on some subset of synonymous
variants is possible, it should be much weaker on average than between missense
mutations, and any interference between selected synonymous variants should
lead to negative LD.

One proposal is that spatial population structure may be responsible for the
increase of LD observed between synonymous variants, as every natural
population lives in some geography that will necessarily induce some structure.
However, neutral evolution alone cannot create this effect. Under standard
population genetic assumptions of constant mutation and recombination rates and
independent mutation events, migration, admixture, and spatial structure do not
result in nonzero $\E[D]$. Rather, for a set of mutations with *given*
differences in allele frequencies between source populations, $D$ can be
nonzero in an admixed population between those particular mutations even when
$D$ is zero in the source populations [@Cavalli-Sforza1971-jv]. But this result
relies on conditioning on directional differences in allele frequencies between
populations. Expected mean allele frequency difference in the source
populations is zero ($\E[(p_1-p_2)(q_1-q_2)] = 0$), as differences are equally
likely to be positive or negative, and resulting $\E[D]$ is still zero when
taking genome-wide averages. Instead, @Sohail2017-zq and @Sandler2020-vk used
forward simulations under multi-population or explicitly spatial models and a
multiplicative fitness function to explore the joint effect of structure,
assortative mating, and interference, scenarios that can lead to nonzero LD.
@Sandler2020-vk find positive LD among neutral mutations in their simulation,
although it extends over much larger distances than observed in the data.
Instead, positive LD in data between synonymous mutations decays to zero within
just a few hundred bases (see Figures \@ref(fig:lddecay) and
\@ref(fig:lddecaydomain), and Figure 5 from @Sandler2020-vk).

@@

When partitioning measurements of LD by mutation classes or regions within
genes, the decreasing number of pairwise comparisons leads to large estimated
measurement noise. Within each population, confidence intervals of observed
$\sigma_d^1$ often overlap with zero or overlap with that of other classes of
mutations. While observed patterns are remarkably consistent across the 15
populations considered here, their joint evolutionary histories make formal
testing of significance difficult due to shared variation, so that cannot be
treated as independent measurements. Extensive forward simulations will likely
be required to more thoroughly assess significance.

From a modeling perspective, in exploring or inferring multi-locus selective
models, the number of plausible selection scenarios becomes quite large as we
relax the strict assumptions of additivity and multiplicative fitness effects.
The inclusion of dominance effects, epistasis, or other interactions leads to a
rapid increase in the number of parameters to consider. This makes performing
forward simulations that span the range of all such selective interaction
scenarios burdensome.

@@

To include selection, we consider a model where we draw lineages uniformly from
the previous generation, but keep lineages with probability proportional to
their fitness. In the absence of dominance, selection reduces to a haploid
model, with acceptance and rejection probabilities depending on the fitnesses
of each haploid copy, where haplotype $Ab$ has fitness $1 + s_{A}$, $aB$ has
fitness $1 + s_{B}$, and $AB$ has fitness $1 + s_{AB}$. We assume the doubly
ancestral haplotype $ab$ has fitness $1$, so fitnesses are relative to that of
$ab$ haplotypes. The standard multiplicative fitness function assumes that
$s_{AB} \approx s_{A} + s_{B}$, and a model for epistasis can be written as
\[s_{AB} = (s_{A} + s_{B}) (1 + \epsilon),\] so that $\epsilon > 0$ implies
synergistic epistasis, while $\epsilon < 0$ implies antagonistic epistasis.

### 2021-03-24 SupportingInformation.Rmd
consolidate figures, incorporate comments
ready for last pass of edits

>>>>

```{r hudson, echo=F, fig.cap="(ref:fighudson)", fig.align="center"}
knitr::include_graphics("figures/hudson_comparison.pdf")
```

(ref:fighudson) The distribution at stationarity of AB haplotype counts in a
sample size of 30, in which we observe 10 A alleles at the left locus, and 10 B
alleles at the right locus.
